Staged Abstract Interpreters

### 2019-09-05 paper/60_cases.tex
ccs
Case Studies

In Section~\ref{sai}, we use a toy language to show that staging an abstract
interpreter is feasible and can be systematically derived from staged concrete
interpreters and unstaged abstract interpreter.
In this section, we conduct several case studies to further show that this
methodology is practically useful and widely applicable to diverse analyses.

@@

We first revisit a similar technique called \textit{abstract compilation} (AC).
AC was introduced by \citet{Boucher:1996:ACN:647473.727587} as an
implementation technique for abstract interpretation based static analysis.
Similar to the present paper, the idea is inspired by partial evaluation:
the program is known statically and the interpretive overhead can be
eliminated. In AC, the compiled analysis can be represented by either text or
closures (higher-order functions). The minute difference is that the closures
can be executed immediately, while the textual programs need to be compiled and
loaded.

@@

In this section, we show that the AC (closure generation as example) can be
understood and implemented as an instance of staging abstract interpreters. We
first revisit the original implementation of abstract compilation of 0-CFA, and
then reproduce their result by simply adding stage annotations.
The generated program of our approach improves approximately the same extent of
speed, but without changing a single line of the analyzer program using
type-based staging annotations. However, AC requires more engineering effort,
i.e., rewriting the whole analyzer into the closure generation form. Moreover,
as shown in Section~\ref{staged_ds}, our approach is able to not only remove the
interpretive overhead, but also enabling several optimizations such specialized
data structures and equivalent rewritings.

@@

\paragraph{Original AC}
The analysis presented by \citeauthor{Boucher:1996:ACN:647473.727587} is 0-CFA
for a small CPS language consisting of abstractions, applications, recursion and
primitive operators. Figure \ref{compare_ac_sai} (left) shows the AC
implementation: different syntactic constructs are handled in different functions
(e.g., @compCall@, @compApp@, etc.).
Before refactoring into AC, these functions would take the store as a
regular argument; now every function returns a value of type @CompAnalysis@,
which is a function (closure) value that takes and returns stores. After the
first run of analysis, we have traversed the AST and obtained a single closure that
takes a store and returns a store. Therefore, the ASTs are static terms, and the
stores are dynamic terms. Now we can invoke the compiled closure with an initial
store to complete the analysis.

@@

On the other side, our approach does exactly the same thing through staging: the
AST are static, and stores are dynamic, therefore we annotate the type @Store@
with @Rep@ (shown in Figure \ref{compare_ac_sai} (right)). In fact, the only
changes of our approach using LMS framework are the types of @Store@ to
@Rep[Store]@, since they will be known at the next stage; other parts do not need to change. 
The generated code
will just looks-up and update the store. The functions at current stage are
eliminated in the residual program, therefore we do not generate higher-order
functions, which provides additional performance improvement. Nevertheless,
the support from LMS framework is necessary, such as staging support for @Map@,
we consider these efforts are reusable and complete orthogonal to the
implementation of analysis.

@@

In this section, to demonstrate the SAI approach is widely applicable, we
extend and refactor the staged abstract interpreter we presented in Section
\ref{sai} to various control-flow analyses (CFA). We first discuss
calling-context sensitive analysis for staged analyzers by tweaking the
allocation strategy. Then we describe how to implement store-widening with
staged monads, which is a common optimization in CFA. Finally, we integrate
abstract garbage collection with staging. All of them are in fact orthogonal to
staging.

@@

The timestamp is essentially a finite list of expressions that tracks the $k$
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to include the identifier and the time when it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //add Time with a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

@@

Every time when we call the @eval@ function, we refresh the timestamp by
calling the @tick@ function, which updates the timestamp in the state monad.
The @tick@ function is implemented as prepending the current control expression
being evaluated (@e@) to the existing calling context, and then taking the
first $k$ elements from the list.  The type @Config@ used in caching is also
changed to include timestamps.

@@

A commonly used technique to improve the running time in control-flow analysis
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, store-widening analysis can be easily
implemented in monadic abstract interpreter by swapping the @StateT@ (for store)
and the @SetT@ (for nondeterminism). In this section, we show this optimization
is orthogonal to staging and we just need to similarly swap the same \textit{staged}
version of monads. After the adjustment, we have the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

Abstract garbage collection is a technique to reclaim unreachable addresses in
the store while doing the abstract interpretation
\cite{Might:2006:IFA:1159803.1159807}. \citet{DBLP:journals/pacmpl/DaraisLNH17}
shows that, for big-step monadic abstract interpreter, we may add a @Reader@
monad to track a set of root addresses and compute the reachable addresses every
time when we obtain a value. Here, we adopt the idea with staged monads:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepReaderT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?],
                 Set[Addr], ?], Env, T] //add a reader monad for Set[Addr]
\end{lstlisting}

@@

\vspace{-1em}
\begin{figure}[h!]
\begin{lstlisting}
  def fix_gc(ev: (Expr => Ans) => (Expr => Ans))(e: Expr): Ans = for {
    ψ <- ask_roots ... // omit code for retrieving ρ, σ, in and out
    val res: Rep[(Set[(Value,Store)], Cache)] = ...  // omit code for checking caches
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1)
    σ <- gc(ψ ⊔ root(vs._2));     _ <- put_store(σ)  // gc performs abstract garbage collection
  } yield v
\end{lstlisting}
\caption{\texttt{fix\_gc} for abstract garbage collection} \label{fixgc}
\end{figure}

@@

\vspace{-2em}
\subsection{Modular Staged Analysis for Free} \label{modular}

One of the challenges of modern static analysis is that programs usually depend
on large libraries and analyzing these libraries is time consuming
\cite{toman_et_al:LIPIcs:2017:7121}.
If we are able to analyze programs and libraries separately and reuse the
result without losing precision, then we can reduce the overhead of repeatedly
analyzing libraries. Some static analyzers compute a summary for a function or
a module, which can be reused later. However, when lacking of the context
information, those analyses can be too conservative or unsound, which both lead
to imprecision. In this section, we show that meta-programming can provide a
trade-off between summary-based modular analysis and slow whole-program
analysis, by utilizing an existing whole-program analyzer.

As we have already seen from Section \ref{sai}, and recall that the primitive
operation @close@ can denote a $\lambda$ term to a next-stage Scala function.
The specializing of the staged abstract interpreter to a $\lambda$ term is
independent to other parts of the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}

@@

We borrow the notation from traditional modular analysis
\cite{DBLP:conf/cc/CousotC02}: such generated function is a modularly reusable
artifact, i.e., independent to its context. 
The generated functions already eliminate the interpretation overhead and
monadic overhead, therefore can be consider as partial summaries. Although, we
still need to run the generated function at next stage to obtain the analysis
result.  At next stage, we can provide different abstract arguments, store and
caches, depending on the context where it is called.

Compared with traditional summary-based analysis, our partial
summaries are represented by executable functions that are
parameterized to the context information. In our approach, the
transition from a whole-program analyzer to modular staged analyzer is
mechanized and for free via staging. We do not need to design any
intermediate representation or data structures of summaries.  In some
scenario, traditional summary-based analyses are able to carry more
information within a module, e.g., the summary of their internal
sub-modules. The staged approach presented here may still need to run
through the fixed-point for sub-modules.

>>>>

In Section~\ref{sai}, we used a toy language to show that staging an abstract
interpreter is feasible and can be systematically derived from staged concrete
interpreters and unstaged abstract interpreters.
In this section, we conduct several case studies to further show that this
methodology is practically useful and widely applicable to diverse analyses.

@@

We first revisit a similar technique called \textit{abstract compilation} (AC).
AC was introduced by \citet{Boucher:1996:ACN:647473.727587} as an
implementation technique for abstract interpretation-based static analysis.
Similar to the present paper, the idea is inspired by partial evaluation:
the program is known statically and the interpretive overhead can be
eliminated. In AC, the compiled analysis can be represented by either text or
closures (higher-order functions). The minute difference is that the closures
can be executed immediately, while the textual programs need to be compiled and
loaded.

@@

In this section, we show that the AC (closure generation as example) can be
understood and implemented as an instance of staging abstract interpreters. We
first revisit the original implementation of abstract compilation of 0-CFA, and
then reproduce their result by simply adding stage annotations.
The program generated by our approach provides approximately the same improvement with regards to
speed, but without changing a single line of the analyzer program using
type-based staging annotations. However, AC requires more engineering effort,
i.e., rewriting the whole analyzer into the closure-generation form. Moreover,
as shown in Section~\ref{staged_ds}, our approach is able to not only remove the
interpretive overhead, but also enable several optimizations such as specialized
data structures and equivalent rewritings.

@@

\paragraph{Original AC}
The analysis presented by \citeauthor{Boucher:1996:ACN:647473.727587} is 0-CFA
for a small CPS language consisting of abstractions, applications, recursion, and
primitive operators. Figure \ref{compare_ac_sai} (left) shows the AC
implementation: different syntactic constructs are handled in different functions
(e.g., @compCall@, @compApp@, etc.).
Before refactoring into AC, these functions would take the store as a
regular argument; now every function returns a value of type @CompAnalysis@,
which is a function (closure) value that takes and returns stores. After the
first run of analysis, we have traversed the AST and obtained a single closure that
takes a store and returns a store. Therefore, the ASTs are static terms, and the
stores are dynamic terms. Now we can invoke the compiled closure with an initial
store to complete the analysis.

@@

On the other side, our approach works equivalently through staging: the
ASTs are static, whereas stores are dynamic; therefore, we annotate the type @Store@
with @Rep@ (shown in Figure \ref{compare_ac_sai} (right)). In fact, the only
changes of our approach using the LMS framework are the types of @Store@ to
@Rep[Store]@, since they will be known at the next stage: no other parts need to change.
The generated code
will consist entirely of looks-up and updating the store. The functions at the current stage are
eliminated in the residual program; therefore, we do not generate higher-order
functions, which provides additional performance improvement. Nevertheless,
the support from the LMS framework is necessary, such as staging support for @Map@.
We consider these efforts reusable and completely orthogonal to the
implementation of the analysis.

@@

In this section, to demonstrate that the SAI approach is widely applicable, we
extend and refactor the staged abstract interpreter we presented in Section
\ref{sai} to various control-flow analyses (CFA). We first discuss
calling-context sensitive analysis for staged analyzers by tweaking the
allocation strategy. Then we describe how to implement store-widening with
staged monads, which is a common optimization in CFA. Finally, we integrate
abstract garbage collection with staging. All of them are in fact orthogonal to
staging.

@@

The timestamp is essentially a finite list of expressions that tracks the $k$-most
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to include an identifier and the time when it gets allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analyses with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //add Time with a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

@@

Every time we call the @eval@ function, we refresh the timestamp by
calling the @tick@ function, which updates the timestamp in the state monad.
The @tick@ function is implemented as prepending the current control expression
being evaluated (@e@) to the existing calling context, and then taking the
first $k$ elements from the list.  The type @Config@ used in caching is also
changed to include timestamps.

@@

A commonly-used technique to improve the running time in control-flow analysis
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, store-widening analysis can be easily
implemented in monadic abstract interpreter by swapping the @StateT@ (for store)
and the @SetT@ (for nondeterminism). In this section, we show this optimization
is orthogonal to staging and we just need to similarly swap the same \textit{staged}
version of monads. After the adjustment, we have the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

Abstract garbage collection is a technique to reclaim unreachable addresses in
the store while doing the abstract interpretation
\cite{Might:2006:IFA:1159803.1159807}. \citet{DBLP:journals/pacmpl/DaraisLNH17}
showed that, for big-step monadic abstract interpreters, we may add a @Reader@
monad to track a set of root addresses and compute the reachable addresses every
time when we obtain a value. Here, we adopt the idea with staged monads:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepReaderT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?],
                 Set[Addr], ?], Env, T] //add a reader monad for Set[Addr]
\end{lstlisting}

@@

\begin{figure}[h!]
\begin{lstlisting}
  def fix_gc(ev: (Expr => Ans) => (Expr => Ans))(e: Expr): Ans = for {
    ψ <- ask_roots ... // omit code for retrieving ρ, σ, in and out
    val res: Rep[(Set[(Value,Store)], Cache)] = ...  // omit code for checking caches
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1)
    σ <- gc(ψ ⊔ root(vs._2));     _ <- put_store(σ)  // gc performs abstract garbage collection
  } yield v
\end{lstlisting}
\caption{\texttt{fix\_gc} for abstract garbage collection} \label{fixgc}
\end{figure}

@@

\subsection{Modular Staged Analysis for Free} \label{modular}

One of the challenges of modern static analysis is that programs usually depend
on large libraries, and analyzing these libraries is time-consuming
\cite{toman_et_al:LIPIcs:2017:7121}.
If we are able to analyze programs and libraries separately and reuse the
results without losing precision, we can reduce the overhead of repeatedly
analyzing libraries. Some static analyzers compute a summary for a function or
a module, which can be reused later. However, when lacking of the context
information, those analyses can be too conservative or unsound, which both lead
to imprecision. In this section, we show that meta-programming can provide a
trade-off between summary-based modular analysis and slow whole-program
analysis by utilizing an existing whole-program analyzer.

As we have already seen from Section \ref{sai}, the primitive
operation @close@ can denote a $\lambda$ term to a next-stage Scala function.
The specializing of the staged abstract interpreter to a $\lambda$ term is
independent to other parts of the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}

@@

We borrow the notation from traditional modular analysis
\cite{DBLP:conf/cc/CousotC02}: such a generated function is a modularly reusable
artifact, i.e., independent to its context.
The generated functions already eliminate the interpretation overhead and
monadic overhead, and can therefore be considered as partial summaries. However, we
still need to run the generated function at next stage to obtain the analysis
result.  At the next stage, we can provide different abstract arguments, stores, and
caches, depending on the context where it is called.

Compared with traditional summary-based analysis, our partial
summaries are represented by executable functions that are
parameterized to the context information. In our approach, the
transition from a whole-program analyzer to a modular-staged analyzer is
mechanized and for-free via staging. We do not need to design any
intermediate representation or data structures of summaries.  In some
scenarios, traditional summary-based analyses are able to carry more
information within a module, e.g., the summary of their internal
submodules. The staged approach presented here may still need to run
through the fixed-point for submodules.

### 2019-09-05 paper/80_related.tex
Case Studies

Splitting an analysis into multiple stages has also been studied for other
forms of static analyses. 
\citet{DBLP:conf/cgo/HardekopfL11} applied staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
programs to obtain a sparse representation, and then the second
stage conducts the flow-sensitive analysis by refining the first
one. Another large body of static analyses are implemented by
generating constraints firstly and then use a solver to discharge the
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Such analyses can be considered having stage distinctions.  One piece of future
work is to explore whether they can be implemented and optimized from
meta-programming. 
Souffl{\'e}, one of the state-of-the-art Datalog engines commonly used in
points-to analysis \cite{Antoniadis:2017:PDS:3088515.3088522}, uses Futamura
projections for efficient implementation \cite{10.1007/978-3-319-41540-6_23}.
However, it only specializes Datalog rules to the running engine, instead of a
full pipeline from a program AST to a specialized solver.

>>>>

Splitting an analysis into multiple stages has also been studied for other
forms of static analyses.
\citet{DBLP:conf/cgo/HardekopfL11} applied staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
programs to obtain a sparse representation, and then the second
stage conducts the flow-sensitive analysis by refining the first
one. Another large body of static analyses are implemented by
first generating constraints and then using a solver to discharge the
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Such analyses can be considered having stage distinctions.  One piece of future
work is to explore whether they can be implemented and optimized from
meta-programming. 
Souffl{\'e}, one of the state-of-the-art Datalog engines commonly used in
points-to analysis \cite{Antoniadis:2017:PDS:3088515.3088522}, uses Futamura
projections for efficient implementation \cite{10.1007/978-3-319-41540-6_23}.
However, it only specializes Datalog rules to the running engine, instead of a
full pipeline from a program AST to a specialized solver.
